Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-19 docs/manuscript.Rmd
updated

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal structure is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three basic competing models for these variables. The causal inference test (CIT) attempts to delineate between them. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation. c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.",
	display=FALSE
)

@@

fig_nums(
	"steiger_sensitivity_plot", 
	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface. When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction. Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y. A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and red surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$. In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$. b) Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative. When $d$ is negative the Steiger test is liable to infer the wrong direction of causality. Shaded regions show the parameter space where $d$ is negative. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.
	",
	display=FALSE
)

fig_nums(
	"cit_mr_comparison_figure", 
	caption="a) Outcomes were simulated to be causally influenced by exposures as shown in the graph, with varying degrees of measurement error applied to both. CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality. The value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality. Rows of graphs represent the sample size used in the simulations. For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality. b) As in (a) except the simulated model was non-causal, and a genetic confounder induces an association between $x$ and $y$. MR is unable to identify this model, so any significant associations are deemed to be incorrect. Outcome 3 denotes evidence for the correct model for the CIT method.",
	display=FALSE
)

@@

In addition, the exposure is influenced by a SNP $g$ such that

@@

Mediation-based analyses that test whether $x$ causally relates to $y$ impinge on evaluating if the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that

@@

The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The methodology of the CIT is as follows. Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ is causally related to an outcome $y$. Thus,

$$
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_g + \beta_g g + \epsilon_g \\
y & = \alpha_x + \beta_x x + \epsilon_x \\
\end{aligned}
$$

The following tests are then performed:

@@

where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome. If all four tests reject the null hypothesis then it is inferred that $x$ is causally related to $y$. The CIT measures the strength of causality by generating an omnibus p-value, $p_{CIT}$, which is simply the largest (least extreme) p-value of the four tests, the intuition being that causal inference is only as strong as the weakest link in the chain of tests.

In these analyses the *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the direction of $x$ causing $y$, and for the direction model of $y$ causing $x$. For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:

- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then the $x \rightarrow y$ model is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the $y \rightarrow x$ model is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no inference is made
- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the confounding model is accepted ($x \leftarrow g \rightarrow y$).

@@

Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample. Assuming the following model

$$
y = \alpha_{MR} + \beta_{MR} \hat{x} + \epsilon_{MR}
$$

where $\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g$, a causal relationship is inferred if the null hypothesis that $\beta_{MR} = 0$ is rejected. A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$ [@Henningsen2007].

@@

- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z > 0$ then a causal association for the correct model is accepted
- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted
- Otherwise no inference is made

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} > \rho_{g,y_o}$. As we show in the appendix, the maximum measurement error of $x_o$ is $\rho_{g,x_o}$ and the minimum is 0 (same applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds. To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement imprecision in $x$ and $y$ we integrate over the simulated $R$ values for the possible measurement error values which are compatible with the direction of causality obtained empirically from the Steiger test. We also do the same for the measurement error values which are not compatible. The ratio of these quantities is an indication of how much of the measurement error space supports the inferred causal direction. A ratio $R=1$ indicates that the Steiger test is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. Full details are provided in Appendix 2.

@@

Two sample MR [@Pierce2013] was performed using the summary statistics presented in [@Shakhbazov2016] for each of `r nrow(shakhbazov)` associations between DNA methylation and gene expression levels. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robust correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Because only summary statistics were available (effect, standard error, effect allele, sample size, p-values) for the instrumental SNP on the methylation and gene expression levels, the Steiger test of two independent correlations was used to infer the direction of causality for each of the associations. The Wald ratio test was then used to estimate the causal effect size for the estimated direction for each association.

@@

Where $z = \rho_{g,y} - \rho_{g,x} < 0$ the measurement error range in $x_o$ and $y_o$ will not change the qualitative inference of the direction of causality. To evaluate the reliability, $R$, of the Steiger test with regards to measurement error, the objective is to compare the weighted proportion of the parameter space that agrees with the inferred direction against the weighted proportion which does not:

@@

If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values. If $R > 1$ then $R$ times as much of the weighted parameter space favours the inferred direction of causality. $V_{z}$, the total volume of the function (`r cfig("steiger_sensitivity_plot")`), can be obtained analytically by solving:

>>>>

fig_nums(
	"dag_figure",
	caption="Gene expression levels (blue blocks) and DNA methylation levels (green triangles) may be correlated but the causal structure is unknown. If a SNP (yellow circle) is associated with both DNA methylation and gene expression levels then it can be used as an instrument, but there are three basic competing models for these variables. The causal inference test (CIT) attempts to distinguish between them. a) Methylation causes gene expression. The left figure shows that the SNP influences methylation levels that in turn influence gene expression levels. The right figure shows the directed acyclic graph that represents this model. Faded symbols represent the measured values whereas solid symbols represent the true values. b) The same as in A, except the causal direction is from gene expression to DNA methylation. c) A model of confounding, where gene expression and DNA methylation are not causally related, but the SNP influences them each through separate pathways or a confounder.",
	display=FALSE
)

@@

fig_nums(
	"steiger_sensitivity_plot", 
	caption="a) We can predict the values the Steiger test would take (z-axis) for different potential values of measurement error (x and y axes), drawn here as the blue surface. When $\\rho_{g,y} > \\rho_{g,x}$, as denoted by the range of values where the blue surface is above the red plane, those values of measurement error lead to our observed Steiger test inferring the wrong causal direction. Where the blue surface lies below the red plane, these measurement error values support the inferred causal direction of X to Y. A measure of reliability, therefore, is the ratio of the negative and positive volumes of the total space bound by the blue and red surfaces, $R = \\frac{V_{z \\geq 0}}{ - V_{z < 0} }$. In this case, where $\\rho_{g,x}^2 = 0.01$ and $\\rho_{x,y}^2 = 0.1$, the $R = 4.40$, which means that 4.40 times as much of the possible measurement error values are in support of the $x \\rightarrow y$ direction of causality than $y \\rightarrow x$. b) Plots depicting the parameter space in which the function $d = cor(x, x_O) - cor(x,y)cor(y, y_O)$ is negative. When $d$ is negative the Steiger test is liable to infer the wrong direction of causality. Shaded regions show the parameter space where $d$ is negative. The graph shows that for the majority of the parameter space of the function, $d$ is positive, especially where causal relationships are relatively weak.
	",
	display=FALSE
)

fig_nums(
	"cit_mr_comparison_figure", 
	caption="a) Outcome $y$ was simulated to be caused by exposure $x$ as shown in the graph, with varying degrees of measurement error applied to both. CIT and MR were used to infer evidence for causality between the exposure and outcome, and to infer the direction of causality. The value of $d = \\rho_{x, x_o} - \\rho_{x,y}\\rho_{y,y_o}$, such that when $d$ is negative we expect the Steiger test to be more likely to be wrong about the direction of causality. Rows of graphs represent the sample size used in the simulations. For the CIT method, outcome 1 denoted evidence for causality with correct model, outcomes 2 or 3 denoted evidence for causality with incorrect model, and outcome 4 denoted no evidence for causality. b) As in (a) except the simulated model was non-causal, and a genetic confounder induces an association between $x$ and $y$. MR is unable to identify this model, so any significant associations are deemed to be incorrect. Outcome 3 denotes evidence for the correct model for the CIT method.",
	display=FALSE
)

@@

In addition, the exposure is influenced by a SNP $g$ with an effect of $\beta_g$ such that

@@

The $\alpha_*$ terms represent intercepts, and the $\epsilon_*$ denote random error. Mediation-based analyses that test whether $x$ causally relates to $y$ impinge on evaluating if the influence of $g$ on $y$ can be accounted for by conditioning on $x$, such that

@@

The CIT method [@Millstein2009] is implemented in the R package *R/cit* [@Millstein2016]. The methodology of the CIT is as follows. Assume an exposure $x$ is instrumented by a SNP $g$, and the exposure $x$ causes an outcome $y$, as described above. The following tests are then performed:

@@

where $y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x$ is the residual of $y$ after adjusting for $x$, where $x$ is assumed to mediate the association between the SNP and the outcome. If all four tests reject the null hypothesis then it is inferred that $x$ causes $y$. The CIT measures the strength of causality by generating an omnibus p-value, $p_{CIT}$, which is simply the largest (least extreme) p-value of the four tests, the intuition being that causal inference is only as strong as the weakest link in the chain of tests.

In these analyses the *cit.cp* function was used to obtain an omnibus p-value. To infer the direction of causality using the CIT method, an omnibus p-value generated by CIT, $p_{CIT}$, was estimated for the direction of $x$ causing $y$ (Model 1), and for the direction of $y$ causing $x$ (Model 2). For some significance threshold $\alpha$, the existence of causality and its direction was inferred based on the following scenarios:

- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then model 1 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then model 2 is accepted
- If $p_{CIT, x \rightarrow y} > \alpha$ and $p_{CIT, y \rightarrow x} > \alpha$ then no evidence for a causal relationship
- If $p_{CIT, x \rightarrow y} < \alpha$ and $p_{CIT, y \rightarrow x} < \alpha$ then the confounding model is accepted ($x \leftarrow g \rightarrow y$).

@@

Two stage least squares (2SLS) is a commonly used technique for performing MR when the exposure, outcome and instrument data are all available in the same sample. A p-value for this test, $p_{MR}$, was obtained using the R package $R/systemfit$ [@Henningsen2007].

@@

- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z > 0$ then a causal association for the correct model is accepted
- If $p_{Steiger} < \alpha$ and $p_{MR} < \alpha$ and $Z < 0$ then a causal association for the incorrect model is accepted
- Otherwise no evidence for a causal relationship

@@

The Steiger test for inferring if $x \rightarrow y$ is based on evaluating $\rho_{gx} > \rho_{gy}$. However, $\rho_{gx}$ and $\rho_{gy}$ are underestimated if $x$ and $y$ are measured imprecisely. If, for example, $x$ has lower measurement precision than $y$ then we might empirically obtain $\rho_{g,x_o} < \rho_{g,y_o}$ because $\rho_{g,x_o}$ could be underestimated more than $\rho_{g,y_o}$.

As we show in Appendix 2 it is possible to infer the bounds of measurement error on $x_o$ or $y_o$ given known genetic associations. The maximum measurement error of $x_o$ is $\rho_{g,x_o}$, because it is known that at least that much of the variance has been explained in $x_o$ by $g$. The minimum is 0, denoting perfectly measured trait values (the same logic applies to $y_o$). It is possible to simulate what the inferred causal direction would be for all values within these bounds.

To evaluate how reliable, $R$, the inference of the causal direction is to potential measurement error in $x$ and $y$ we need to predict the values of $\rho_{gy} - \rho_{gx}$ for those values of measurement error. We integrate over the entire range of $\rho_{gy} - \rho_{gx}$ values for possible measurement error values. We find the ratio of the volume that agrees with the inferred direction of causality over the volume that disagrees with the inferred direction of causality. A ratio $R=1$ indicates that the inferred causal direction is highly sensitive to measurement error, with equal weight of the measurement error parameter space supporting both directions of causality. In general, the $R$ value denotes that the inferred direction of causality is $R$ times more likely to be the empirical result than the opposite direction. Full details are provided in Appendix 2.

@@

Two sample MR [@Pierce2013] was performed using the summary statistics for genetic influences on gene expression and DNA methylation. To do this we obtained a list of `r nrow(qtl_orig)` gene expression - DNA methylation associations as reported in Shakhbazov et al [-@Shakhbazov2016]. These were filtered to be located on the same chromosome, have robust correlations after correcting for multiple testing, and to share a SNP that had a robust cis-acting effect on both the DNA methylation probe and the gene expression probe. Because only summary statistics were available (effect, standard error, effect allele, sample size, p-values) for the instrumental SNP on the methylation and gene expression levels, the Steiger test of two independent correlations was used to infer the direction of causality for each of the associations. The Wald ratio test was then used to estimate the causal effect size for the estimated direction for each association.

@@

Call $z = \rho_{g,y} - \rho_{g,x}$ the true difference in the variance explained by the genetic variant in $y$ and $x$. If $z < 0$ then we infer that $x \rightarrow y$. There will be some values of $\rho_{x,x_o}$ and $\rho_{y,y_o}$ that do not alter whether $z < 0$. To evaluate the reliability, $R$, of the inference of the causal direction with regards to measurement error, the objective is to compare the  proportion of the parameter space that agrees with the inferred direction against the proportion which does not:

@@

If $R=1$ then the direction of causality is equally probable across the range of possible measurement error values. If $R > 1$ then $R$ times as much of the parameter space favours the inferred direction of causality. $V_{z}$, the total volume of the function (`r cfig("steiger_sensitivity_plot")`), can be obtained analytically by solving:

### 2017-02-19 docs/manuscript.tex
updated

In addition, the exposure is influenced by a SNP \(g\) such that

@@

Mediation-based analyses that test whether \(x\) causally relates to
\(y\) impinge on evaluating if the influence of \(g\) on \(y\) can be
accounted for by conditioning on \(x\), such that

@@

The CIT method\textsuperscript{4} is implemented in the R package
\emph{R/cit}\textsuperscript{18}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) is causally related to an outcome \(y\). Thus,

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
x & = \alpha_g + \beta_g g + \epsilon_g \\
y & = \alpha_x + \beta_x x + \epsilon_x \\
\end{aligned}
\]

The following tests are then performed:

@@

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
mediate the association between the SNP and the outcome. If all four
tests reject the null hypothesis then it is inferred that \(x\) is
causally related to \(y\). The CIT measures the strength of causality by
generating an omnibus p-value, \(p_{CIT}\), which is simply the largest
(least extreme) p-value of the four tests, the intuition being that
causal inference is only as strong as the weakest link in the chain of
tests.

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
for the direction of \(x\) causing \(y\), and for the direction model of
\(y\) causing \(x\). For some significance threshold \(\alpha\), the
existence of causality and its direction was inferred based on the
following scenarios:

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then the \(x \rightarrow y\)
  model is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then the \(y \rightarrow x\)
  model is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no inference is made
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then the confounding model is
  accepted (\(x \leftarrow g \rightarrow y\)).
\end{itemize}

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. Assuming the following model

\[
y = \alpha_{MR} + \beta_{MR} \hat{x} + \epsilon_{MR}
\]

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{37}.

@@

\begin{itemize}
\tightlist
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
  a causal association for the correct model is accepted
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
  a causal association for the incorrect model is accepted
\item
  Otherwise no inference is made
\end{itemize}

@@

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} > \rho_{g,y_o}\).
As we show in the appendix, the maximum measurement error of \(x_o\) is
\(\rho_{g,x_o}\) and the minimum is 0 (same applies to \(y_o\)). It is
possible to simulate what the inferred causal direction would be for all
values within these bounds. To evaluate how reliable, \(R\), the
inference of the causal direction is to potential measurement
imprecision in \(x\) and \(y\) we integrate over the simulated \(R\)
values for the possible measurement error values which are compatible
with the direction of causality obtained empirically from the Steiger
test. We also do the same for the measurement error values which are not
compatible. The ratio of these quantities is an indication of how much
of the measurement error space supports the inferred causal direction. A
ratio \(R=1\) indicates that the Steiger test is highly sensitive to
measurement error, with equal weight of the measurement error parameter
space supporting both directions of causality. Full details are provided
in Appendix 2.

@@

Two sample MR\textsuperscript{28} was performed using the summary
statistics presented in\textsuperscript{40} for each of 4122
associations between DNA methylation and gene expression levels. To do
this we obtained a list of 458 gene expression - DNA methylation
associations as reported in Shakhbazov et al\textsuperscript{40}. These
were filtered to be located on the same chromosome, have robust
correlations after correcting for multiple testing, and to share a SNP
that had a robust cis-acting effect on both the DNA methylation probe
and the gene expression probe. Because only summary statistics were
available (effect, standard error, effect allele, sample size, p-values)
for the instrumental SNP on the methylation and gene expression levels,
the Steiger test of two independent correlations was used to infer the
direction of causality for each of the associations. The Wald ratio test
was then used to estimate the causal effect size for the estimated
direction for each association.

@@

Figure 1: Gene expression levels (blue blocks) and DNA methylation
levels (green triangles) may be correlated but the causal structure is
unknown. If a SNP (yellow circle) is associated with both DNA
methylation and gene expression levels then it can be used as an
instrument, but there are three basic competing models for these
variables. The causal inference test (CIT) attempts to delineate between
them. a) Methylation causes gene expression. The left figure shows that
the SNP influences methylation levels that in turn influence gene
expression levels. The right figure shows the directed acyclic graph
that represents this model. Faded symbols represent the measured values
whereas solid symbols represent the true values. b) The same as in A,
except the causal direction is from gene expression to DNA methylation.
c) A model of confounding, where gene expression and DNA methylation are
not causally related, but the SNP influences them each through separate
pathways or a confounder.

@@

Figure 4: a) We can predict the values the Steiger test would take
(z-axis) for different potential values of measurement error (x and y
axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
as denoted by range of values where the blue surface is above the red
plane, those values of measurement error lead to our observed Steiger
test inferring the wrong causal direction. Where the blue surface lies
below the red plane, these measurement error values support the inferred
causal direction of X to Y. A measure of reliability, therefore, is the
ratio of the negative and positive volumes of the total space bound by
the blue and red surfaces, \(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\).
In this case, where \(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\),
the \(R = 4.40\), which means that 4.40 times as much of the possible
measurement error values are in support of the \(x \rightarrow y\)
direction of causality than \(y \rightarrow x\). b) Plots depicting the
parameter space in which the function
\(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is negative. When \(d\) is
negative the Steiger test is liable to infer the wrong direction of
causality. Shaded regions show the parameter space where \(d\) is
negative. The graph shows that for the majority of the parameter space
of the function, \(d\) is positive, especially where causal
relationships are relatively weak.

@@

Figure 5: a) Outcomes were simulated to be causally influenced by
exposures as shown in the graph, with varying degrees of measurement
error applied to both. CIT and MR were used to infer evidence for
causality between the exposure and outcome, and to infer the direction
of causality. The value of
\(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when \(d\) is
negative we expect the Steiger test to be more likely to be wrong about
the direction of causality. Rows of graphs represent the sample size
used in the simulations. For the CIT method, outcome 1 denoted evidence
for causality with correct model, outcomes 2 or 3 denoted evidence for
causality with incorrect model, and outcome 4 denoted no evidence for
causality. b) As in (a) except the simulated model was non-causal, and a
genetic confounder induces an association between \(x\) and \(y\). MR is
unable to identify this model, so any significant associations are
deemed to be incorrect. Outcome 3 denotes evidence for the correct model
for the CIT method.

@@

Where \(z = \rho_{g,y} - \rho_{g,x} < 0\) the measurement error range in
\(x_o\) and \(y_o\) will not change the qualitative inference of the
direction of causality. To evaluate the reliability, \(R\), of the
Steiger test with regards to measurement error, the objective is to
compare the weighted proportion of the parameter space that agrees with
the inferred direction against the weighted proportion which does not:

@@

If \(R=1\) then the direction of causality is equally probable across
the range of possible measurement error values. If \(R > 1\) then \(R\)
times as much of the weighted parameter space favours the inferred
direction of causality. \(V_{z}\), the total volume of the function
(Figure 4), can be obtained analytically by solving:

>>>>

In addition, the exposure is influenced by a SNP \(g\) with an effect of
\(\beta_g\) such that

@@

The \(\alpha_*\) terms represent intercepts, and the \(\epsilon_*\)
denote random error. Mediation-based analyses that test whether \(x\)
causally relates to \(y\) impinge on evaluating if the influence of
\(g\) on \(y\) can be accounted for by conditioning on \(x\), such that

@@

The CIT method\textsuperscript{4} is implemented in the R package
\emph{R/cit}\textsuperscript{18}. The methodology of the CIT is as
follows. Assume an exposure \(x\) is instrumented by a SNP \(g\), and
the exposure \(x\) causes an outcome \(y\), as described above. The
following tests are then performed:

@@

where \(y - \hat{y} = y - \hat{\alpha}_g + \hat{\beta}_g x\) is the
residual of \(y\) after adjusting for \(x\), where \(x\) is assumed to
mediate the association between the SNP and the outcome. If all four
tests reject the null hypothesis then it is inferred that \(x\) causes
\(y\). The CIT measures the strength of causality by generating an
omnibus p-value, \(p_{CIT}\), which is simply the largest (least
extreme) p-value of the four tests, the intuition being that causal
inference is only as strong as the weakest link in the chain of tests.

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
for the direction of \(x\) causing \(y\) (Model 1), and for the
direction of \(y\) causing \(x\) (Model 2). For some significance
threshold \(\alpha\), the existence of causality and its direction was
inferred based on the following scenarios:

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then model 1 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then model 2 is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no evidence for a causal
  relationship
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then the confounding model is
  accepted (\(x \leftarrow g \rightarrow y\)).
\end{itemize}

@@

Two stage least squares (2SLS) is a commonly used technique for
performing MR when the exposure, outcome and instrument data are all
available in the same sample. A p-value for this test, \(p_{MR}\), was
obtained using the R package \(R/systemfit\)\textsuperscript{37}.

@@

\begin{itemize}
\tightlist
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z > 0\) then
  a causal association for the correct model is accepted
\item
  If \(p_{Steiger} < \alpha\) and \(p_{MR} < \alpha\) and \(Z < 0\) then
  a causal association for the incorrect model is accepted
\item
  Otherwise no evidence for a causal relationship
\end{itemize}

@@

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} < \rho_{g,y_o}\)
because \(\rho_{g,x_o}\) could be underestimated more than
\(\rho_{g,y_o}\).

As we show in Appendix 2 it is possible to infer the bounds of
measurement error on \(x_o\) or \(y_o\) given known genetic
associations. The maximum measurement error of \(x_o\) is
\(\rho_{g,x_o}\), because it is known that at least that much of the
variance has been explained in \(x_o\) by \(g\). The minimum is 0,
denoting perfectly measured trait values (the same logic applies to
\(y_o\)). It is possible to simulate what the inferred causal direction
would be for all values within these bounds.

To evaluate how reliable, \(R\), the inference of the causal direction
is to potential measurement error in \(x\) and \(y\) we need to predict
the values of \(\rho_{gy} - \rho_{gx}\) for those values of measurement
error. We integrate over the entire range of \(\rho_{gy} - \rho_{gx}\)
values for possible measurement error values. We find the ratio of the
volume that agrees with the inferred direction of causality over the
volume that disagrees with the inferred direction of causality. A ratio
\(R=1\) indicates that the inferred causal direction is highly sensitive
to measurement error, with equal weight of the measurement error
parameter space supporting both directions of causality. In general, the
\(R\) value denotes that the inferred direction of causality is \(R\)
times more likely to be the empirical result than the opposite
direction. Full details are provided in Appendix 2.

@@

Two sample MR\textsuperscript{28} was performed using the summary
statistics for genetic influences on gene expression and DNA
methylation. To do this we obtained a list of 458 gene expression - DNA
methylation associations as reported in Shakhbazov et
al\textsuperscript{40}. These were filtered to be located on the same
chromosome, have robust correlations after correcting for multiple
testing, and to share a SNP that had a robust cis-acting effect on both
the DNA methylation probe and the gene expression probe. Because only
summary statistics were available (effect, standard error, effect
allele, sample size, p-values) for the instrumental SNP on the
methylation and gene expression levels, the Steiger test of two
independent correlations was used to infer the direction of causality
for each of the associations. The Wald ratio test was then used to
estimate the causal effect size for the estimated direction for each
association.

@@

Figure 1: Gene expression levels (blue blocks) and DNA methylation
levels (green triangles) may be correlated but the causal structure is
unknown. If a SNP (yellow circle) is associated with both DNA
methylation and gene expression levels then it can be used as an
instrument, but there are three basic competing models for these
variables. The causal inference test (CIT) attempts to distinguish
between them. a) Methylation causes gene expression. The left figure
shows that the SNP influences methylation levels that in turn influence
gene expression levels. The right figure shows the directed acyclic
graph that represents this model. Faded symbols represent the measured
values whereas solid symbols represent the true values. b) The same as
in A, except the causal direction is from gene expression to DNA
methylation. c) A model of confounding, where gene expression and DNA
methylation are not causally related, but the SNP influences them each
through separate pathways or a confounder.

@@

Figure 4: a) We can predict the values the Steiger test would take
(z-axis) for different potential values of measurement error (x and y
axes), drawn here as the blue surface. When \(\rho_{g,y} > \rho_{g,x}\),
as denoted by the range of values where the blue surface is above the
red plane, those values of measurement error lead to our observed
Steiger test inferring the wrong causal direction. Where the blue
surface lies below the red plane, these measurement error values support
the inferred causal direction of X to Y. A measure of reliability,
therefore, is the ratio of the negative and positive volumes of the
total space bound by the blue and red surfaces,
\(R = \frac{V_{z \geq 0}}{ - V_{z < 0} }\). In this case, where
\(\rho_{g,x}^2 = 0.01\) and \(\rho_{x,y}^2 = 0.1\), the \(R = 4.40\),
which means that 4.40 times as much of the possible measurement error
values are in support of the \(x \rightarrow y\) direction of causality
than \(y \rightarrow x\). b) Plots depicting the parameter space in
which the function \(d = cor(x, x_O) - cor(x,y)cor(y, y_O)\) is
negative. When \(d\) is negative the Steiger test is liable to infer the
wrong direction of causality. Shaded regions show the parameter space
where \(d\) is negative. The graph shows that for the majority of the
parameter space of the function, \(d\) is positive, especially where
causal relationships are relatively weak.

@@

Figure 5: a) Outcome \(y\) was simulated to be caused by exposure \(x\)
as shown in the graph, with varying degrees of measurement error applied
to both. CIT and MR were used to infer evidence for causality between
the exposure and outcome, and to infer the direction of causality. The
value of \(d = \rho_{x, x_o} - \rho_{x,y}\rho_{y,y_o}\), such that when
\(d\) is negative we expect the Steiger test to be more likely to be
wrong about the direction of causality. Rows of graphs represent the
sample size used in the simulations. For the CIT method, outcome 1
denoted evidence for causality with correct model, outcomes 2 or 3
denoted evidence for causality with incorrect model, and outcome 4
denoted no evidence for causality. b) As in (a) except the simulated
model was non-causal, and a genetic confounder induces an association
between \(x\) and \(y\). MR is unable to identify this model, so any
significant associations are deemed to be incorrect. Outcome 3 denotes
evidence for the correct model for the CIT method.

@@

Call \(z = \rho_{g,y} - \rho_{g,x}\) the true difference in the variance
explained by the genetic variant in \(y\) and \(x\). If \(z < 0\) then
we infer that \(x \rightarrow y\). There will be some values of
\(\rho_{x,x_o}\) and \(\rho_{y,y_o}\) that do not alter whether
\(z < 0\). To evaluate the reliability, \(R\), of the inference of the
causal direction with regards to measurement error, the objective is to
compare the proportion of the parameter space that agrees with the
inferred direction against the proportion which does not:

@@

If \(R=1\) then the direction of causality is equally probable across
the range of possible measurement error values. If \(R > 1\) then \(R\)
times as much of the parameter space favours the inferred direction of
causality. \(V_{z}\), the total volume of the function (Figure 4), can
be obtained analytically by solving:
